## A few R data analysis demonstrations

### From the 4-Feb-2016 LOPSA-NJ Meeting - Updated 8-Feb-2016

This RMarkdown + knitr version allows Markdown content outside of the R code sections.

R code sections have "echo=TRUE, message=TRUE, warning=TRUE" to show all work.

One way to generate HTML is with a command like:

    Rscript -e "setwd('/n2/r-reports/'); library(knitr); knit2html('~/lab/R/quickStart1.Rmd')"

setwd sets the directory where the intermediate .md and final .html files are written. The argument to knit2html is the path to the "source code".

RStudio has an interactive GUI way to do the same thing

## Initialize

```{r initialize, echo=TRUE, message=TRUE, warning=TRUE}
## First R code block

## Load common libraries in a good order, note the function overloading warnings

library(ggplot2)
library(plyr)  # plyr, followed by dplyr
library(dplyr) # Provides %>% and functions used for filtering and aggregation - See RStudio's Data Wrangling Cheatsheet
library(readr)
library(lubridate)

Sys.setenv(TZ="UTC") # Often prevents TZ issues

FigureWidth  <- 16
FigureHeight <- 8

## A custom ggplot2 theme that is decent for HTML reports
theme_jm2 <- theme_bw() +
    theme(
        plot.title  = element_text(size = rel(1.5), family = 'Helvetica', face = 'bold'),
        axis.title  = element_text(size = rel(1.5), colour = "black", face = 'bold'),
        axis.text.x = element_text(size = rel(1.5), lineheight = 0.9, colour = "black", vjust = 1, face = 'bold'),
        axis.text.y = element_text(size = rel(1.5), lineheight = 0.9, colour = "black", hjust = 1, face = 'bold'),
        legend.text = element_text(size = rel(1.3)),
        plot.title = element_text(lineheight=.8, face="bold", size = 10),
        plot.margin = unit(c(0.0, 1, -0.6, 0.5), "lines"),
        strip.text.y = element_text(size = rel(1.7), colour = "black", face = 'bold') # Enhance readability of facet labels
    )

```

## Example 1 - Small Number of Data Points - Inline

```{r Example1, echo=TRUE, message=TRUE, fig.width = FigureWidth, fig.height = FigureHeight}
##```{r Example1, echo=FALSE, message=FALSE, fig.width = FigureWidth, fig.height = FigureHeight}
## swapping the two lines above will turn off the code and results echo and only the plot will appear in the output

## Note that read_delim from the readr package requires a single space between fields

IPSraw <- 'Time tcp_syn.dropped tcp_syn.forwarded
2015-09-17T21:33 49050 48483
2015-09-17T21:36 87309 85551
2015-09-17T21:39 163092 99578
2015-09-17T21:42 247875 114235
2015-09-17T21:45 335129 129098
2015-09-17T21:48 405430 135635
2015-09-17T21:51 473972 143137
2015-09-17T21:54 541985 149912
2015-09-17T21:57 651037 158139
2015-09-17T22:00 759434 166562
2015-09-17T22:03 812341 170244
2015-09-17T22:06 877437 174047
2015-09-17T22:09 942562 177693
2015-09-17T22:12 987158 180936'

IPS <- read_delim(IPSraw, delim = ' ', col_types = list(Time = col_datetime('%Y-%m-%dT%H:%M')))

PointSize <- 1.0

Title <- 'Server to Backend - SYNs Dropped / 3 Minutes'

ggplot(IPS) +
    geom_line(aes(x = Time, y = tcp_syn.dropped), size=0.1) +
    geom_point(aes(x = Time, y = tcp_syn.dropped),
    size=PointSize + 2, color = 'red', shape=19) +
    xlab('') +
    ggtitle(Title) + theme_jm2

## Have a Look at the Data, three ways

IPS           # Just dump the data

str(IPS)      # Data structure information

glimpse(IPS)  # A dplyr function

```

## Example 2 - Data Wrangling

```{r Example2, echo=TRUE, message=TRUE, fig.width = FigureWidth, fig.height = FigureHeight}

site_servers <- list() # Servers of interest
site_servers[["nyc"]] <- c("nyc-bc-drp-01", "nyc-bc-drp-02")   # A form of hash
site_servers[["bos"]] <- c("bos-bc-drp-01", "bos-bc-drp-02")

servers_of_interest <- unique(unlist(site_servers))            # Vector of server names

File <- 'si20151117.dat'                                       ## If R can find the file without a path ##
File <- '/Users/meekj/wpl/talks/lopsanj-2016-1/si20151117.dat' ## Otherwise, adjust the path for your data location ##
File <- '/home/meekj/wpl/talks/lopsanj-2016-1/si20151117.dat'  ## Otherwise, adjust the path for your data location ##

bc <- read.table(File, header = TRUE)                          # Base R read.table
bc <- bc %>% filter(Server %in% servers_of_interest)           # Keep only servers of interest
bc$Time <- as.POSIXct(bc$utime, tz="UTC", origin="1970-01-01") # Convert UNIX seconds to POSIXct

global_agg <- bc %>% group_by(Time) %>% summarise(TotalUsers = sum(Users), ServerCount = n()) # Pipe-like

MedianServerCount <- median(global_agg$ServerCount)                   # How many servers are active?
global_agg <- global_agg %>% filter(ServerCount == MedianServerCount) # Drop minutes with missing server data

bc$Egress  <- substr(bc$Server, 1, 3)            # Get data center site code from server name
SiteCodes  <- unique(bc$Egress)                  # Character vector of unique DC names
cSiteCodes <- paste(SiteCodes, collapse = ' & ') # Collapse into simple string

PointSize <- 1.0

Title <- paste('Concurrent Users via', cSiteCodes, 'Using', MedianServerCount, 'Servers')

ggplot(global_agg) +
    geom_line(aes(x = Time, y = TotalUsers), size=PointSize - 0.6, color = 'dodgerblue') + # Line first
    geom_point(aes(x = Time, y = TotalUsers), size=PointSize, color = 'blue', shape=19) +  # Points on top
    xlab("") + ylab("Global Concurrent Users") +
    scale_x_datetime(date_minor_breaks = "1 hour") + # ggplot2 2.0.0, remove 'date_' for previous versions!
    ggtitle(Title) + theme_jm2

## Look at data structures

str(bc)

str(global_agg)

PointSize <- 0.6

```

```{r Example2b, echo=TRUE, message=TRUE, fig.width = FigureWidth, fig.height = 3*FigureHeight}

## Faceted plot, separate for each server, note 3*FigureHeight above to make the plot taller to accommodate the four faceted plots

ggplot(bc) +
    geom_point(aes(x = Time, y = Users), size=PointSize, color = 'blue', shape=19) +  # Points on top
    xlab("") + ylab("Concurrent Users") +
    scale_x_datetime(date_minor_breaks = "1 hour") + # ggplot2 2.0.0, remove 'date_' for previous versions!
    facet_grid(Server ~ ., scales="free_y") +
    ggtitle(Title) + theme_jm2
#    theme(strip.text.y = element_text(size = rel(1.2), colour = "black", face = 'bold')) # Enhance readability of facet labels

```

```{r Example2c, echo=TRUE, message=TRUE, fig.width = FigureWidth, fig.height = FigureHeight}

## Single plot, each server gets a different color

ggplot(bc) +
    geom_point(aes(x = Time, y = Users, color = Server), size=PointSize, shape=19) +  # Points on top
    xlab("") + ylab("Concurrent Users") +
    scale_x_datetime(date_minor_breaks = "1 hour") + # ggplot2 2.0.0, remove 'date_' for previous versions!
    ggtitle(Title) +
    scale_colour_manual(values=c("blue", "red", "green", "yellow")) + theme_jm2


```
## Example 3 - More Data Wrangling and Boxplots

```{r Example3, echo=TRUE, message=TRUE, fig.width = FigureWidth, fig.height = FigureHeight}

## Re-read the data to get all the server data again
#
File <- 'si20151117.dat'                                       ## If R can find the file without a path ##
File <- '/Users/meekj/wpl/talks/lopsanj-2016-1/si20151117.dat' ## Otherwise, adjust the path for your data location ##
File <- '/home/meekj/wpl/talks/lopsanj-2016-1/si20151117.dat'  ## Otherwise, adjust the path for your data location ##

bc <- read.table(File, header = TRUE)                          # Base R read.table

## Look only at servers with '-sg-' in their name
#
all_servers <- as.character(unique(bc$Server))                 # vector of all server names
servers_of_interest <- all_servers[grep('-sg-', all_servers)]  # grep returns indicies

bc <- bc %>% filter(Server %in% servers_of_interest)           # Keep only servers of interest
bc$Time <- as.POSIXct(bc$utime, tz="UTC", origin="1970-01-01") # Convert UNIX seconds to POSIXct

Title <- 'CPU Utilization'

## Make a boxplot with actual data points displayed with geom_jitter
## Boxplot goes on top of data points, otherwise it would be obscurred
## Set outlier.colour=NA so that redundant outlier points are not plotted
## alpha sets the transparency of the box so that the 'jittered' data points can be seen

ggplot(bc) +
  geom_jitter(aes(group = Server, x = Server, y = CPU), colour  = "dodgerblue", size = 0.8, shape=19) +
  geom_boxplot(aes(group = Server, x = Server, y = CPU), fill = "darkblue", outlier.colour=NA, alpha=0.2) +
  xlab('') + ylab('') +
  ggtitle(Title) + theme_jm2 + theme(axis.text.x=element_text(angle = 60, hjust=1))

Title <- 'CPU Utilization per 100 Users'
bc$CPUper100Users <- 100 * bc$CPU / bc$Users # Generate a new  data frame column in the 'Base R way', compare to dplyr::mutate used below

ggplot(bc) +
  geom_jitter(aes(group = Server, x = Server, y = CPUper100Users), colour  = "dodgerblue", size = 0.8, shape=19) +
  geom_boxplot(aes(group = Server, x = Server, y = CPUper100Users), fill = "lightblue", outlier.colour=NA, alpha=0.2) +
  xlab('') + ylab('') +
  ggtitle(Title) + theme_jm2 + theme(axis.text.x=element_text(angle = 60, hjust=1))

## CPU / User is less meaningful at low user counts (or low CPU utilization) so we will set a threshold level for the next plot

## Compute 95th percentile Concurrent User count for each server of interest
server_summary <- bc %>% group_by(Server) %>% summarise(q95Users = quantile(Users, probs=0.95))

server_summary # Look at summary

ThresholdPercent <- 30 # Threshold in percent

## Compute reporting threshold by adding a column to the data frame
server_summary <- server_summary %>% mutate(Threshold = ThresholdPercent * q95Users / 100)

server_summary$Server <- as.character(server_summary$Server) # Change data type from factor to character string

server_summary # Look at summary again

## Make a new data frame dropping points below the threshold for each server
bct <- NULL # Start empty
for (server in servers_of_interest) { # Build data frame one server at a time (using a somewhat ugly way to reference the threshold value)
    bct <- rbind(bct, bc %>% filter(Server == server & Users >= server_summary[server_summary$Server == server, ]$Threshold))
}

## Check that the filtering worked

bc  %>% group_by(Server) %>% summarise(Count = n(), MinUsers = min(Users)) # Full data set
bct %>% group_by(Server) %>% summarise(Count = n(), MinUsers = min(Users)) # Reduced data set

Title <- paste('CPU Utilization per 100 Users Above ', ThresholdPercent, '% of 95th Percentile per Server', sep = '')

ggplot(bct) +
  geom_jitter(aes(group = Server, x = Server, y = CPUper100Users), colour  = "dodgerblue", size = 0.4, shape=19) +
  geom_boxplot(aes(group = Server, x = Server, y = CPUper100Users), fill = "lightblue", outlier.colour=NA, alpha=0.4) +
  xlab('') + ylab('') +
  ggtitle(Title) + theme_jm2 + theme(axis.text.x=element_text(angle = 60, hjust=1))

```
***
##### Jon Meek, PhD - meekj (at) ieee.org - February 2016

